On the combination of Set Functions and Failures

### 2013-02-14 docs/src/manual.tex
Manual updated

\lstnewenvironment{curry}{}{}
\lstnewenvironment{currynomath}{\lstset{mathescape=false}}{} % Curry w/o math

\newcommand{\listline}{\vrule width0pt depth1.75ex}

@@

\newcommand{\labs}{\mbox{\tt\char92}}  % lambda abstraction in Curry
\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
\newcommand{\cb}{CurryBrowser\xspace}

@@

This document describes KiCS2 (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
Version \textbf{2}),
an implementation of the multi-paradigm language Curry
\cite{Hanus97POPL,Hanus06Curry} that is
based on compiling Curry programs into Haskell programs.
Curry is a universal programming language aiming at the amalgamation
of the most important declarative programming paradigms,
namely functional programming and logic programming.  
Curry combines in a seamless way features from functional programming
(nested expressions, lazy evaluation, higher-order functions),
logic programming (logical variables, partial data structures,
built-in search), and concurrent programming (concurrent evaluation
of constraints with synchronization on logical variables).
The current KiCS2 implementation does not support concurrent
constraints. Alternatively, one can write distributed applications
by the use of sockets that can be registered and accessed
with symbolic names.
Moreover, KiCS2 also supports
the high-level implementation of
graphical user interfaces and web services
(as described in more detail in
\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).

We assume familiarity with the ideas and features
of Curry as described in the Curry language definition \cite{Hanus12Curry}.
Therefore, this document only explains the use of the different
components of KiCS2
and the differences and restrictions of KiCS2
(see Section~\ref{sec-restrictions})
compared with the language Curry (Version 0.8.3).
The basic ideas of the implementation of KiCS2
can be found in
\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.

@@

\section{Overview of KiCS2}

@@

This version of KiCS2 has been developed and
tested on Linux systems.
In principle, it should be also executable on other
platforms on which a Haskell implementation (Glasgow Haskell Compiler
and Cabal) exists, like in many Linux distributions, Sun Solaris,
or Mac OS X systems.

Installation instructions for KiCS2 can be found in
the file \code{INSTALL.txt} stored in the KiCS2 installation directory.
Note that there are two possibilities to install KiCS2:
\begin{description}
\item[Global installation:]\index{global installation}\index{installation!global}
KiCS2 is installed in some global system directory where users
have no write permission. In this case, some options for experimenting
with KiCS2 (like \code{supply} or \code{ghc}, see below)
are not available (since they require the recompilation of parts of
the installed system).
\item[Local installation:]\index{local installation}\index{installation!local}
KiCS2 is installed in some local user directory where the user
has write permission and the option \code{GLOBALINSTALL}
in the \code{Makefile} of the KiCS2 installation is set as follows:
\begin{lstlisting}
GLOBALINSTALL=no
\end{lstlisting}
In this case, all options of KiCS2 are available.
\end{description}
In the following, $kics2home$ denotes the installation directory
of the KiCS2 installation.

@@

All executables required to use the different components
of KiCS2 are stored in the directory \code{$kics2home$/bin}.
You should add this directory
to your path (e.g., by the \code{bash} command
\ccode{export PATH=$kics2home$/bin:\$PATH}).

@@

Since the translation of Curry programs with KiCS2 creates
some auxiliary files (see Section~\ref{sec-auxfiles} for details),
you need write permission
in the directory where you have stored your Curry programs.
Moreover, the current implementation also recompiles
system libraries according to the setting of some options.
Therefore, the KiCS2 system should be locally installed
in your user account.
The auxiliary files for all Curry programs in the current
directory can be deleted by the command\pindex{cleancurry}
\begin{curry}
cleancurry
\end{curry}
(this is a shell script stored in the \code{bin} directory of the
KiCS2 installation, see above).
The command
\begin{curry}
cleancurry -r
\end{curry}
also deletes the auxiliary files in all subdirectories.

@@

There are a few minor restrictions on Curry programs
when they are processed with KiCS2:
\begin{itemize}
\item
\index{singleton variables}\index{variables!singleton}
\emph{Singleton pattern variables}, i.e., variables that occur only once
in a rule, should be denoted as an anonymous variable \ccode{_},
otherwise the parser will print a warning since this is a
typical source of programming errors.
\item
KiCS2 translates all \emph{local declarations} into global functions with
additional arguments (``lambda lifting'', see Appendix~D of the
Curry language report).
Thus, in the various run-time systems, the definition of
functions with local declarations look different from
their original definition (in order to see the result
of this transformation, you can use the \cb, see
Section~\ref{sec-currybrowser}).
\item \index{tabulator stops}
Tabulator stops instead of blank spaces in source files are
interpreted as stops at columns 9, 17, 25, 33, and so on.
In general, tabulator stops should be avoided in source programs.
\item
Encapsulated search\index{encapsulated search}: In order
to allow the integration of non-deterministic computations
in programs performing I/O at the top-level, KiCS2 provides
the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
and \code{SearchTree} (Section~\ref{Library:SearchTree})
where one can defined search operators and various search strategies.
The general definition of encapsulated search of the Curry report
\cite{HanusSteiner98PLILP} is not supported.
\item Concurrent computations\index{concurrency}
based on the suspension of expressions
containing free variables are not yet supported.
KiCS2 supports \emph{value generators} for free variables
so that a free variable is instantiated when its value is demanded.
For instance, the initial expression
\begin{curry}
x == True where x free
\end{curry}
is non-deterministically evaluated to \code{False} and \code{True}
by instantiating \code{x} to \code{False} and \code{True}, respectively.
Thus, a computation is never suspended due to free variables.
This behavior also applies to free variables of primitive types
like integers. For instance, the initial expression
\begin{curry}
x*y=:=1 where x,y free
\end{curry}
is non-deterministically evaluated to the two solutions
\begin{curry}
{x = -1, y = -1} Success
{x = 1, y = 1} Success
\end{curry}
\item
Unification is performed without an occur check.
\item
There is currently no general connection to external constraint solvers.
\end{itemize}

\subsection{Modules in KiCS2}
\label{sec-modules}

The current implementation of KiCS2 supports only flat module names,
i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
In order to allow the structuring of modules in different directories,
KiCS2 searches for imported modules in various directories.
By default, imported modules are searched in the directory
of the main program and the system module directories
\ccode{$kics2home$/lib} and \ccode{$kics2home$/lib/meta}.
This search path can be extended
by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
(which can be also set in a KiCS2 session by the option
\ccode{:set path}\pindex{path}\pindex{:set path},
see below)
to a list of directory names separated by colons (\ccode{:}).
In addition, a local standard search path
can be defined in the \ccode{.kics2rc} file
(see Section~\ref{sec-customization}).
Thus, modules to be loaded are searched in the following
directories (in this order, i.e., the first occurrence of a module file
in this search path is imported):
\begin{enumerate}
\item Current working directory (\ccode{.}) or directory prefix
of the main module (e.g., directory \ccode{/home/joe/curryprogs}
if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
\item The directories enumerated in the environment variable \code{CURRYPATH}.
\item The directories enumerated in the \ccode{.kics2rc} variable
      \ccode{libraries}.
\item The directories \ccode{$kics2home$/lib} and \ccode{$kics2home$/lib/meta}.
\end{enumerate}
Note that the standard prelude (\code{$kics2home$/lib/Prelude.curry})
will be always implicitly imported to all modules if a module
does not contain an explicit import declaration for the module
\code{Prelude}.

@@

\section{Using the Interactive Environment of KiCS2}
\label{sec-kics2}

This section describes the interactive environment
KiCS2\index{KiCS2}
that supports the development of applications written in Curry.
The implementation of KiCS2 contains also a separate compiler
which is automatically invoked by the interactive environment.

\subsection{Invoking KiCS2}
\label{sec:invoke-kics2}

To start KiCS2, execute the command
\ccode{kics2}\pindex{kics2}
(this is a shell script stored in
\code{$kics2home$/bin} where $kics2home$ is the installation directory
of KiCS2).
When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
the prelude (\code{$kics2home$/lib/Prelude.curry})
is already loaded, i.e., all definitions in the prelude are accessible.
Now you can type various commands (see next section)
or an expression to be evaluated.

One can also invoke KiCS2 with parameters.
These parameters are usual a sequence of commands
(see next section) that are executed before the user interaction
starts. For instance, the invocation
\begin{curry}
kics2 :load Mod :add List
\end{curry}
starts KiCS2, loads the main module \code{Mod}, and adds the additional
module \code{List}.
The invocation
\begin{curry}
kics2 :load Mod :eval config
\end{curry}
starts KiCS2, loads the main module \code{Mod}, and evaluates
the operation \code{config} before the user interaction starts.
As a final example, the invocation
\begin{curry}
kics2 :load Mod :save :quit
\end{curry}
starts KiCS2, loads the main module \code{Mod}, creates an executable,
and terminates KiCS2. This invocation could be useful in ``make'' files
for systems implemented in Curry.

\subsection{Command of KiCS2}
\label{sec:kics2-commands}

The {\bf most important commands} of KiCS2 are
(it is sufficient to type a unique prefix of a command if it is unique,
e.g., one can type \ccode{:r} instead of \ccode{:reload}):

@@

If the free variables in the initial goal are of a polymorphic type,
as in the expression
\begin{curry}
xs++ys=:=[z] where xs,ys,z free
\end{curry}
they are specialized to the type \ccode{()}
(since the current implementation of KiCS2 does not support
computations with polymorphic logic variables).

@@

\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
Similar to \ccode{:interface}
but shows the interface of the module $m$
which must be in the load path of KiCS2.

@@

\item[\fbox{\code{:set $option$}}]\pindex{:set}
Set or turn on/off a specific option
of the KiCS2 environment (see \ref{sec:options} for a description
of all options). Options are turned on by the prefix
\ccode{+} and off by the prefix \ccode{-}. Options that can only
be set (e.g., \code{path}) must not contain a prefix.

@@

\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
The expression $expr$, which is typically of type \ccode{IO ()},
is evaluated in an independent process which runs in
parallel to the current KiCS2 process.
All output and error messages from this new process are suppressed.
This command is useful to test distributed Curry programs
where one can start
a new server process by this command. The new process
will be terminated when the evaluation of the expression $expr$
is finished.

@@

\subsection{Option of KiCS2}
\label{sec:options}

@@

\begin{description}
\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
for loading modules to $path$.
Note that this search path is only used for loading modules
inside this invocation of KiCS2.

@@

\item[\fbox{\code{choices $n$}}]\pindex{choices}
Show the internal choice structure
(according to the implementation described in
\cite{BrasselHanusPeemoellerReck11})
resulting from the complete evaluation of the main expression
in a tree-like structure.
This mode is only useful for debugging or
understanding the implementation of non-deterministic
evaluations used in KiCS2.
If the optional argument $n$ is provided,
the tree is shown up to depth $n$.

@@

\item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
Define additional options passed to the KiCS2 compiler.
For instance, setting the option
\begin{curry}
:set cmp -O 0
\end{curry}
has the effect that all optimizations performed by the KiCS2 compiler
are turned off.

\item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
Define additional options passed to the Glasgow Haskell Compiler (GHC)
when the generated Haskell programs are compiled.
One has to be careful when using such options.
For instance, in a global installation of KiCS2
(see Section~\ref{sec-install}), libraries are pre-compiled
so that inconsistencies might occur if compilation options
might be changed.

@@

If the evaluation of operations in some main module loaded into
KiCS2 requires specific options, like an iterative-deepening
search strategy, one can also put these options into the source
code of this module in order to avoid setting these options
every time when this module is loaded.
Such
{\bf source-file options}\index{source-file option}\index{option!in source file}
must occur before the module header, i.e., before the first declaration
(module header, imports, fixity declaration, defining rules, etc)
occurring in the module.
Each source file option must be in a line of the form
\begin{curry}
{-# KiCS2_OPTION $opt$ #-}
\end{curry}
where $opt$ is an option that can occur in a \ccode{:set} command
(compare Section~\ref{sec:options}).
Such a line in the source code (which is a comment according to
the syntax of Curry)
has the effect that this option is set by the KiCS2 command
\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
as a main module. For instance, if a module starts with the
lines
\begin{curry}
{-# KiCS2_OPTION ids #-}
{-# KiCS2_OPTION +ghci #-}
{-# KiCS2_OPTION v2 #-}
module M where
$\ldots$
\end{curry}
then the load command \ccode{:load M} will also
set the options for iterative deepening, using \code{ghci}
and verbosity level 2.

@@

In order to have support for line editing or history functionality
in the command line of KiCS2 (as often supported by the \code{readline}
library), you should have the Unix command \code{rlwrap} installed
on your local machine.
If \code{rlwrap} is installed, it is used by KiCS2 if called on a terminal.
If it should not be used (e.g., because it is executed
in an editor with \code{readline} functionality), one can
call KiCS2 with the parameter \ccode{--noreadline}
(which must occur as the first parameter).

@@

In order to customize the behavior of KiCS2 to your own preferences,
there is a configuration file which is read by KiCS2 when it is invoked.
When you start KiCS2 for the first time, a standard version of
this configuration file is copied with the name
\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
into your home directory. The file contains definitions
of various settings, e.g., about showing warnings, using Curry extensions,
programs etc.
After you have started KiCS2 for the first time, look into this file
and adapt it to your own preferences.

@@

Emacs is a powerful programmable editor suitable for program development.
It is freely available for many platforms
(see \url{http://www.emacs.org}).
The distribution of KiCS2 contains also a special
\emph{Curry mode}\index{Curry mode}\index{Emacs}
that supports the development of Curry programs in
the Emacs environment.
This mode includes support for syntax highlighting,
finding declarations in the current buffer, and
loading Curry programs into KiCS2
in an Emacs shell.

The Curry mode has been adapted from a similar mode for Haskell programs.
Its installation is described in the file \code{README}
in directory \ccode{$kics2home$/tools/emacs} which also contains
the sources of the Curry mode and a short description about
the use of this mode.

@@

KiCS2 supports some extensions in Curry programs that are not (yet)
part of the definition of Curry. These extensions are described below.

@@

Local variable declarations (introduced by \code{let}\pindex{let}
or \code{where}\pindex{where}) can be (mutually) recursive in KiCS2.
For instance, the declaration
\begin{curry}
ones5 = let ones = 1 : ones
         in take 5 ones
\end{curry}
introduces the local variable \code{ones} which is bound
to a \emph{cyclic structure}\index{cyclic structure}
representing an infinite list of \code{1}'s.
Similarly, the definition
\begin{curry}
onetwo n = take n one2
 where
   one2 = 1 : two1
   two1 = 2 : one2
\end{curry}
introduces a local variables \code{one2} that represents
an infinite list of alternating \code{1}'s and \code{2}'s
so that the expression \code{(onetwo 6)} evaluates to \code{[1,2,1,2,1,2]}.

@@

In contrast to the basic concept in \cite{Leijen05}, KiCS2/Curry provides a
simpler version of records. Some of the features described there are
currently not supported or even restricted.

@@

\cb is part of the KiCS2 distribution and can be
started in two ways:
\begin{itemize}
\item
In the command shell via the command: \code{$kics2home$/bin/currybrowser mod}
\item
In the KiCS2 environment after loading the module
\code{mod} and typing the command \ccode{:browse}.
\end{itemize}
Here, \ccode{mod} is the name of the main module of a Curry application.
After the start, \cb loads the interfaces of the main
module and all imported modules before a GUI is created
for interactive browsing.

@@

CurryTest\index{CurryTest}\index{testing programs}\index{program!testing}
is a simple tool in the KiCS2 distribution to write
and run repeatable tests. CurryTest simplifies the task
of writing test cases for a module and executing them.
The tool is easy to use. Assume one has implemented a module \code{MyMod}
and wants to write some test cases to test its functionality,
making regression tests in future versions, etc.
For this purpose, there is a system library \code{Assertion}
(Section~\ref{Library:Assertion}) which
contains the necessary definitions for writing tests.
In particular, it exports an abstract polymorphic type \ccode{Assertion a}
together with the following operations:
\begin{curry}
assertTrue      :: String -> Bool -> Assertion ()
assertEqual     :: String -> a -> a -> Assertion a
assertValues    :: String -> a -> [a] -> Assertion a
assertSolutions :: String -> (a->Success) -> [a] -> Assertion a
assertIO        :: String -> IO a -> a -> Assertion a
assertEqualIO   :: String -> IO a -> IO a -> Assertion a
\end{curry}
The expression \ccode{assertTrue $s$ $b$}
is an assertion (named $s$) that the expression $b$ has the value \code{True}.
Similarly, the expression \ccode{assertEqual $s$ $e_1$ $e_2$}
asserts that the expressions $e_1$ and $e_2$
must be equal (i.e., \code{$e_1$==$e_2$} must hold),
the expression \ccode{assertValues $s$ $e$ $vs$} asserts
that $vs$ is the multiset of all values of $e$,
and the expression \ccode{assertSolutions $s$ $c$ $vs$} asserts
that the constraint abstraction $c$ has the multiset of solutions $vs$.
Furthermore, the expression \ccode{assertIO $s$ $a$ $v$}
asserts that the I/O action $a$ yields the value $v$ whenever it is
executed, and
the expression \ccode{assertEqualIO $s$ $a_1$ $a_2$}
asserts that the I/O actions $a_1$ and $a_2$ yields equal values.
The name $s$ provided as a first argument in each assertion
is used in the protocol produced by the test tool.

@@

\section{Libraries of the KiCS2 Distribution}
\label{sec:libraries}

>>>>

\lstnewenvironment{curry}{}{}
\lstnewenvironment{currynomath}{\lstset{mathescape=false}}{} % Curry w/o math
\newcommand{\listline}{\vrule width0pt depth1.75ex}

@@

\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
\newcommand{\CYS}{KiCS2\xspace} % name of Curry system described here
\newcommand{\cyshome}{\mbox{\textit{kics2home}}\xspace} % symbolic installation directory
\newcommand{\cb}{CurryBrowser\xspace}

@@

This document describes \CYS (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
Version \textbf{2}),
an implementation of the multi-paradigm language Curry
\cite{Hanus97POPL,Hanus06Curry} that is
based on compiling Curry programs into Haskell programs.
Curry is a universal programming language aiming at the amalgamation
of the most important declarative programming paradigms,
namely functional programming and logic programming.  
Curry combines in a seamless way features from functional programming
(nested expressions, lazy evaluation, higher-order functions),
logic programming (logical variables, partial data structures,
built-in search), and concurrent programming (concurrent evaluation
of constraints with synchronization on logical variables).
The current \CYS implementation does not support concurrent
constraints. Alternatively, one can write distributed applications
by the use of sockets that can be registered and accessed
with symbolic names.
Moreover, \CYS also supports
the high-level implementation of
graphical user interfaces and web services
(as described in more detail in
\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).

We assume familiarity with the ideas and features
of Curry as described in the Curry language definition \cite{Hanus12Curry}.
Therefore, this document only explains the use of the different
components of \CYS
and the differences and restrictions of \CYS
(see Section~\ref{sec-restrictions})
compared with the language Curry (Version 0.8.3).
The basic ideas of the implementation of \CYS
can be found in
\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.

@@

\section{Overview of \CYS}

@@

This version of \CYS has been developed and
tested on Linux systems.
In principle, it should be also executable on other
platforms on which a Haskell implementation (Glasgow Haskell Compiler
and Cabal) exists, like in many Linux distributions, Sun Solaris,
or Mac OS X systems.

Installation instructions for \CYS can be found in
the file \code{INSTALL.txt} stored in the \CYS installation directory.
Note that there are two possibilities to install \CYS:
\begin{description}
\item[Global installation:]\index{global installation}\index{installation!global}
\CYS is installed in some global system directory where users
have no write permission. In this case, some options for experimenting
with \CYS (like \code{supply} or \code{ghc}, see below)
are not available (since they require the recompilation of parts of
the installed system).
\item[Local installation:]\index{local installation}\index{installation!local}
\CYS is installed in some local user directory where the user
has write permission and the option \code{GLOBALINSTALL}
in the \code{Makefile} of the \CYS installation is set as follows:
\begin{lstlisting}
GLOBALINSTALL=no
\end{lstlisting}
In this case, all options of \CYS are available.
\end{description}
In the following, \cyshome denotes the installation directory
of the \CYS installation.

@@

All executables required to use the different components
of \CYS are stored in the directory \code{\cyshome/bin}.
You should add this directory
to your path (e.g., by the \code{bash} command
\ccode{export PATH=\cyshome/bin:\$PATH}).

@@

Since the translation of Curry programs with \CYS creates
some auxiliary files (see Section~\ref{sec-auxfiles} for details),
you need write permission
in the directory where you have stored your Curry programs.
Moreover, the current implementation also recompiles
system libraries according to the setting of some options.
Therefore, the \CYS system should be locally installed
in your user account.
The auxiliary files for all Curry programs in the current
directory can be deleted by the command\pindex{cleancurry}
\begin{curry}
cleancurry
\end{curry}
(this is a shell script stored in the \code{bin} directory of the
\CYS installation, see above).
The command
\begin{curry}
cleancurry -r
\end{curry}
also deletes the auxiliary files in all subdirectories.

@@

There are a few minor restrictions on Curry programs
when they are processed with \CYS:
\begin{itemize}
\item
\index{singleton variables}\index{variables!singleton}
\emph{Singleton pattern variables}, i.e., variables that occur only once
in a rule, should be denoted as an anonymous variable \ccode{_},
otherwise the parser will print a warning since this is a
typical source of programming errors.
\item
\CYS translates all \emph{local declarations} into global functions with
additional arguments (``lambda lifting'', see Appendix~D of the
Curry language report).
Thus, in the various run-time systems, the definition of
functions with local declarations look different from
their original definition (in order to see the result
of this transformation, you can use the \cb, see
Section~\ref{sec-currybrowser}).
\item \index{tabulator stops}
Tabulator stops instead of blank spaces in source files are
interpreted as stops at columns 9, 17, 25, 33, and so on.
In general, tabulator stops should be avoided in source programs.
\item
Encapsulated search\index{encapsulated search}: In order
to allow the integration of non-deterministic computations
in programs performing I/O at the top-level, \CYS provides
the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
and \code{SearchTree} (Section~\ref{Library:SearchTree})
where one can defined search operators and various search strategies.
The general definition of encapsulated search of the Curry report
\cite{HanusSteiner98PLILP} is not supported.
\item Concurrent computations\index{concurrency}
based on the suspension of expressions
containing free variables are not yet supported.
\CYS supports \emph{value generators} for free variables
so that a free variable is instantiated when its value is demanded.
For instance, the initial expression
\begin{curry}
x == True where x free
\end{curry}
is non-deterministically evaluated to \code{False} and \code{True}
by instantiating \code{x} to \code{False} and \code{True}, respectively.
Thus, a computation is never suspended due to free variables.
This behavior also applies to free variables of primitive types
like integers. For instance, the initial expression
\begin{curry}
x*y=:=1 where x,y free
\end{curry}
is non-deterministically evaluated to the two solutions
\begin{curry}
{x = -1, y = -1} Success
{x = 1, y = 1} Success
\end{curry}
\item
Unification is performed without an occur check.
\item
There is currently no general connection to external constraint solvers.
\end{itemize}

\subsection{Modules in \CYS}
\label{sec-modules}

The current implementation of \CYS supports only flat module names,
i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
In order to allow the structuring of modules in different directories,
\CYS searches for imported modules in various directories.
By default, imported modules are searched in the directory
of the main program and the system module directories
\ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
This search path can be extended
by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
(which can be also set in a \CYS session by the option
\ccode{:set path}\pindex{path}\pindex{:set path},
see below)
to a list of directory names separated by colons (\ccode{:}).
In addition, a local standard search path
can be defined in the \ccode{.kics2rc} file
(see Section~\ref{sec-customization}).
Thus, modules to be loaded are searched in the following
directories (in this order, i.e., the first occurrence of a module file
in this search path is imported):
\begin{enumerate}
\item Current working directory (\ccode{.}) or directory prefix
of the main module (e.g., directory \ccode{/home/joe/curryprogs}
if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
\item The directories enumerated in the environment variable \code{CURRYPATH}.
\item The directories enumerated in the \ccode{.kics2rc} variable
      \ccode{libraries}.
\item The directories \ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
\end{enumerate}
Note that the standard prelude (\code{\cyshome/lib/Prelude.curry})
will be always implicitly imported to all modules if a module
does not contain an explicit import declaration for the module
\code{Prelude}.

@@

\section{Using the Interactive Environment of \CYS}
\label{sec-kics2}

This section describes the interactive environment
\CYS\index{\CYS}
that supports the development of applications written in Curry.
The implementation of \CYS contains also a separate compiler
which is automatically invoked by the interactive environment.

\subsection{Invoking \CYS}
\label{sec:invoke-kics2}

To start \CYS, execute the command
\ccode{kics2}\pindex{kics2}
(this is a shell script stored in
\code{\cyshome/bin} where \cyshome is the installation directory
of \CYS).
When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
the prelude (\code{\cyshome/lib/Prelude.curry})
is already loaded, i.e., all definitions in the prelude are accessible.
Now you can type various commands (see next section)
or an expression to be evaluated.

One can also invoke \CYS with parameters.
These parameters are usual a sequence of commands
(see next section) that are executed before the user interaction
starts. For instance, the invocation
\begin{curry}
kics2 :load Mod :add List
\end{curry}
starts \CYS, loads the main module \code{Mod}, and adds the additional
module \code{List}.
The invocation
\begin{curry}
kics2 :load Mod :eval config
\end{curry}
starts \CYS, loads the main module \code{Mod}, and evaluates
the operation \code{config} before the user interaction starts.
As a final example, the invocation
\begin{curry}
kics2 :load Mod :save :quit
\end{curry}
starts \CYS, loads the main module \code{Mod}, creates an executable,
and terminates \CYS. This invocation could be useful in ``make'' files
for systems implemented in Curry.

\subsection{Command of \CYS}
\label{sec:kics2-commands}

The {\bf most important commands} of \CYS are
(it is sufficient to type a unique prefix of a command if it is unique,
e.g., one can type \ccode{:r} instead of \ccode{:reload}):

@@

If the free variables in the initial goal are of a polymorphic type,
as in the expression
\begin{curry}
xs++ys=:=[z] where xs,ys,z free
\end{curry}
they are specialized to the type \ccode{()}
(since the current implementation of \CYS does not support
computations with polymorphic logic variables).

@@

\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
Similar to \ccode{:interface}
but shows the interface of the module $m$
which must be in the load path of \CYS.

@@

\item[\fbox{\code{:set $option$}}]\pindex{:set}
Set or turn on/off a specific option
of the \CYS environment (see \ref{sec:options} for a description
of all options). Options are turned on by the prefix
\ccode{+} and off by the prefix \ccode{-}. Options that can only
be set (e.g., \code{path}) must not contain a prefix.

@@

\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
The expression $expr$, which is typically of type \ccode{IO ()},
is evaluated in an independent process which runs in
parallel to the current \CYS process.
All output and error messages from this new process are suppressed.
This command is useful to test distributed Curry programs
where one can start
a new server process by this command. The new process
will be terminated when the evaluation of the expression $expr$
is finished.

@@

\subsection{Options of \CYS}
\label{sec:options}

@@

\begin{description}
\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
for loading modules to $path$.
Note that this search path is only used for loading modules
inside this invocation of \CYS.

@@

\item[\fbox{\code{choices $n$}}]\pindex{choices}
Show the internal choice structure
(according to the implementation described in
\cite{BrasselHanusPeemoellerReck11})
resulting from the complete evaluation of the main expression
in a tree-like structure.
This mode is only useful for debugging or
understanding the implementation of non-deterministic
evaluations used in \CYS.
If the optional argument $n$ is provided,
the tree is shown up to depth $n$.

@@

\item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
Define additional options passed to the \CYS compiler.
For instance, setting the option
\begin{curry}
:set cmp -O 0
\end{curry}
has the effect that all optimizations performed by the \CYS compiler
are turned off.

\item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
Define additional options passed to the Glasgow Haskell Compiler (GHC)
when the generated Haskell programs are compiled.
One has to be careful when using such options.
For instance, in a global installation of \CYS
(see Section~\ref{sec-install}), libraries are pre-compiled
so that inconsistencies might occur if compilation options
might be changed.

@@

If the evaluation of operations in some main module loaded into
\CYS requires specific options, like an iterative-deepening
search strategy, one can also put these options into the source
code of this module in order to avoid setting these options
every time when this module is loaded.
Such
{\bf source-file options}\index{source-file option}\index{option!in source file}
must occur before the module header, i.e., before the first declaration
(module header, imports, fixity declaration, defining rules, etc)
occurring in the module.
Each source file option must be in a line of the form
\begin{curry}
{-# KiCS2_OPTION $opt$ #-}
\end{curry}
where $opt$ is an option that can occur in a \ccode{:set} command
(compare Section~\ref{sec:options}).
Such a line in the source code (which is a comment according to
the syntax of Curry)
has the effect that this option is set by the \CYS command
\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
as a main module. For instance, if a module starts with the
lines
\begin{curry}
{-# KiCS2_OPTION ids #-}
{-# KiCS2_OPTION +ghci #-}
{-# KiCS2_OPTION v2 #-}
module M where
$\ldots$
\end{curry}
then the load command \ccode{:load M} will also
set the options for iterative deepening, using \code{ghci}
and verbosity level 2.

@@

In order to have support for line editing or history functionality
in the command line of \CYS (as often supported by the \code{readline}
library), you should have the Unix command \code{rlwrap} installed
on your local machine.
If \code{rlwrap} is installed, it is used by \CYS if called on a terminal.
If it should not be used (e.g., because it is executed
in an editor with \code{readline} functionality), one can
call \CYS with the parameter \ccode{--noreadline}
(which must occur as the first parameter).

@@

In order to customize the behavior of \CYS to your own preferences,
there is a configuration file which is read by \CYS when it is invoked.
When you start \CYS for the first time, a standard version of
this configuration file is copied with the name
\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
into your home directory. The file contains definitions
of various settings, e.g., about showing warnings, using Curry extensions,
programs etc.
After you have started \CYS for the first time, look into this file
and adapt it to your own preferences.

@@

Emacs is a powerful programmable editor suitable for program development.
It is freely available for many platforms
(see \url{http://www.emacs.org}).
The distribution of \CYS contains also a special
\emph{Curry mode}\index{Curry mode}\index{Emacs}
that supports the development of Curry programs in
the Emacs environment.
This mode includes support for syntax highlighting,
finding declarations in the current buffer, and
loading Curry programs into \CYS
in an Emacs shell.

The Curry mode has been adapted from a similar mode for Haskell programs.
Its installation is described in the file \code{README}
in directory \ccode{\cyshome/tools/emacs} which also contains
the sources of the Curry mode and a short description about
the use of this mode.

@@

\CYS supports some extensions in Curry programs that are not (yet)
part of the definition of Curry. These extensions are described below.

@@

Local variable declarations (introduced by \code{let}\pindex{let}
or \code{where}\pindex{where}) can be (mutually) recursive in \CYS.
For instance, the declaration
\begin{curry}
ones5 = let ones = 1 : ones
         in take 5 ones
\end{curry}
introduces the local variable \code{ones} which is bound
to a \emph{cyclic structure}\index{cyclic structure}
representing an infinite list of \code{1}'s.
Similarly, the definition
\begin{curry}
onetwo n = take n one2
 where
   one2 = 1 : two1
   two1 = 2 : one2
\end{curry}
introduces a local variables \code{one2} that represents
an infinite list of alternating \code{1}'s and \code{2}'s
so that the expression \code{(onetwo 6)} evaluates to \code{[1,2,1,2,1,2]}.

@@

In contrast to the basic concept in \cite{Leijen05}, \CYS/Curry provides a
simpler version of records. Some of the features described there are
currently not supported or even restricted.

@@

\cb is part of the \CYS distribution and can be
started in two ways:
\begin{itemize}
\item
In the command shell via the command: \code{\cyshome/bin/currybrowser mod}
\item
In the \CYS environment after loading the module
\code{mod} and typing the command \ccode{:browse}.
\end{itemize}
Here, \ccode{mod} is the name of the main module of a Curry application.
After the start, \cb loads the interfaces of the main
module and all imported modules before a GUI is created
for interactive browsing.

@@

CurryTest\index{CurryTest}\index{testing programs}\index{program!testing}
is a simple tool in the \CYS distribution to write
and run repeatable tests. CurryTest simplifies the task
of writing test cases for a module and executing them.
The tool is easy to use. Assume one has implemented a module \code{MyMod}
and wants to write some test cases to test its functionality,
making regression tests in future versions, etc.
For this purpose, there is a system library \code{Assertion}
(Section~\ref{Library:Assertion}) which
contains the necessary definitions for writing tests.
In particular, it exports an abstract polymorphic type \ccode{Assertion a}
together with the following operations:
\begin{curry}
assertTrue      :: String -> Bool -> Assertion ()
assertEqual     :: String -> a -> a -> Assertion a
assertValues    :: String -> a -> [a] -> Assertion a
assertSolutions :: String -> (a->Success) -> [a] -> Assertion a
assertIO        :: String -> IO a -> a -> Assertion a
assertEqualIO   :: String -> IO a -> IO a -> Assertion a
\end{curry}
The expression \ccode{assertTrue $s$ $b$}
is an assertion (named $s$) that the expression $b$ has the value \code{True}.
Similarly, the expression \ccode{assertEqual $s$ $e_1$ $e_2$}
asserts that the expressions $e_1$ and $e_2$
must be equal (i.e., \code{$e_1$==$e_2$} must hold),
the expression \ccode{assertValues $s$ $e$ $vs$} asserts
that $vs$ is the multiset of all values of $e$,
and the expression \ccode{assertSolutions $s$ $c$ $vs$} asserts
that the constraint abstraction $c$ has the multiset of solutions $vs$.
Furthermore, the expression \ccode{assertIO $s$ $a$ $v$}
asserts that the I/O action $a$ yields the value $v$ whenever it is
executed, and
the expression \ccode{assertEqualIO $s$ $a_1$ $a_2$}
asserts that the I/O actions $a_1$ and $a_2$ yields equal values.
The name $s$ provided as a first argument in each assertion
is used in the protocol produced by the test tool.

@@

\section{Libraries of the \CYS Distribution}
\label{sec:libraries}
